---
title: "Genexpi evaluation"
output: html_notebook
---

This R notebook performs the evaluation of Genexpi against TD-Aracne on regulon identifiaction tasks. Note that the full evaluation of Genexpi can take hours and that the evaluation of TD-Aracne can take several days on an high end desktop.

# Setting up

```{r setup}
  options(java.parameters = "-Xmx2048m")
```


First we install and start all the required libraries.

```{r}
  # Helper function to install packages if they are not available
  pkgTest <- function(x)
  {
    if (!require(x,character.only = TRUE))
    {
      install.packages(x,dep=TRUE)
        if(!require(x,character.only = TRUE)) stop("Package not found")
    }
  }

  # Install and load Genexpi from GitHub vie devtools
  pkgTest("devtools")
  install_github("cas-bioinf/genexpi", subdir="rpackage")

  # Install and load the packages required for evaluation against TD-Aracne
  source("https://bioconductor.org/biocLite.R")
  biocLite(c("Biobase","TDARACNE", "RBGL"))
  library(Biobase)
  library(TDARACNE)
  library(RBGL)
  
  pkgTest("foreach")
  library(foreach)
  pkgTest("doParallel")
  library(doParallel)


```


Genexpi relies on OpenCL, so we need to check the available OpenCL devices:

```{r}
listOpenCLDevices()
```
If the list is empty or the call ends in an error, it is probably because your computer has no OpenCL driver installed. Almost all recent GPUs and CPUs (processors) support OpenCL, so you should be able to run Genexpi.

* For GPUs, try updating your device driver. We also encountered cases where GPUs started being available only after installing a CPU driver (see next bullet).
* For CPUs (the usual processors), you need to install specific drivers. As of 2017-05-22, drivers for Intel CPUs can be downloaded at https://software.intel.com/en-us/articles/opencl-drivers#latest_CPU_runtime - (you want “runtime-only”). AMD’s drivers can be found at http://support.amd.com/en-us/kb-articles/Pages/OpenCL2-Driver.aspx.

If the list is non-empty, you can choose which device to use for computation. The best devices to use with Genexpi are (in the following order):

* A dedicated graphics card (GPU) for computing (not connected to a display)
* A Xeon Phi card
* Your processor (CPU)
* A GPU connected to a display

While GPUs will run Genexpi the fastest, using a GPU connected to a display is discouraged as it interferes with the operating system (OS) and may cause your computer to freeze and/or the computations to be reset by the operating system. If you absolutely need to use the GPU running your display, and use Windows, you may want to disable [TDR - Timeout detection and recovery](https://msdn.microsoft.com/en-us/library/windows/hardware/ff569918(v=vs.85).aspx) (expert only).

With that in mind, let us select a device.

```{r}
#Select the device automatically
#deviceSpecs = getDeviceSpecs()

#Select the best GPU device
#deviceSpecs = getDeviceSpecs(deviceType = "gpu")

#Select the best CPU device
#deviceSpecs = getDeviceSpecs(deviceType = "processor")

#Select device by ID
deviceSpecs = getDeviceSpecs(device = 1)

#Display the selected device
if(is.null(deviceSpecs)) {
  stop("Invalid device")
} else {
  paste0("Using device: ",deviceSpecs$getDevice()$toString())
}
```
# Testing SigB regulon of Bacilus subtilis

## Load the sigB data from GSE6865

The original source: (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE6865)

```{r}
temp <- tempfile();
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE6nnn/GSE6865/matrix/GSE6865_series_matrix.txt.gz", temp)
gse6865_raw_df = read.delim(gzfile(temp), comment.char = "!") #Intermediate data frame representation

# Raw profile data. We scale by 1000 to get to more wieldy values
gse6865_raw = as.matrix(gse6865_raw_df[,2:15]) / 1000
rownames(gse6865_raw) = gse6865_raw_df$ID_REF

#Times (in minutes) for the individual samples
gse6865_raw_time = c(0,5,10,15,20,25,30,40,50,60,70,80,90,100)
colnames(gse6865_raw) <- sapply(gse6865_raw_time,  FUN = function(x) { paste0(x,"min")})

#We will compute at 1 minute resolution
smoothTime = 0:100
errorDef = list(relative = 0.2, absolute = 0, minimal = 0.5)

# There are a few genes that have NA values for the first or second measurement
cat("Numbers of NA values per measurement (NAs will be imputed):\n")
apply(gse6865_raw, MARGIN  = 2, FUN = function(x) { sum(is.na(x)) })

# For the first measurement, we can expect the value to be 0 (the series are from germination)
gse6865_raw[is.na(gse6865_raw[,1]),1] = 0

# For the second measurement we will go with a linear interpolation
na2 = is.na(gse6865_raw[,2])
gse6865_raw[na2,2] = 0.5 * gse6865_raw[na2,1] + 0.5 * gse6865_raw[na2,3]



# Genes in the regulon according to http://subtiwiki.uni-goettingen.de/wiki/index.php/SigB_regulon as of 2017-02-09
sigBRegulonNames = c("aag", "aldY", "bmr", "bmrR", "bmrU", "bofC", "cdd", "chaA", "clpC", "clpP", "corA", "csbA", "csbB", "csbC", "csbD", "csbX", "ctc", "ctsR", "cypC", "disA", "dps", "era", "gabD", "gsiB", "gspA", "gtaB", "guaD", "hpf", "iolW", "ispD", "ispF", "katE", "katX", "mcsA", "mcsB", "mgsR", "mhqO", "mhqP", "nadE", "nhaX", "ohrB", "opuE", "phoP", "phoR", "plsC", "radA", "rnr", "rpmEB", "rsbRD", "rsbV", "rsbW", "rsbX",  "sodA", "spo0E", "spoVC", "spx", "trxA", "yaaH", "yaaI", "yacL", "ybyB", "ycbP", "ycdF", "ycdG", "yceC", "yceD", "yceE", "yceF", "yceG", "yceH", "ydaD", "ydaE", "ydaF", "ydaG", "ydaJ", "ydaK", "ydaL", "ydaM", "ydaN", "ydaP", "ydaS", "ydaT", "ydbD", "ydeC", "ydhK", "yerD", "yfhD", "yfhE", "yfhF", "yfhK", "yfhL", "yfhM", "yfkD", "yfkH", "yfkI", "yfkJ", "yfkM", "yfkS", "yfkT", "yflA", "yflH", "yflT", "ygxB", "yhcM", "yhdF", "yhdN", "yhxD", "yitT", "yjbC", "yjgB", "yjgC", "yjgD", "yjzE", "ykgA", "ykgB", "ykuT", "ykzI", "ylxP", "ymzB", "yoaA", "yocB", "yocK", "yoxB", "yoxC", "ypuB", "ypuC", "ypuD", "yqhB", "yqhP", "yqhQ", "yqjL", "yraA", "ysdB", "ysnF", "ytaB", "ytkL", "ytxG", "ytxH", "ytxJ", "yugU", "yuzA", "yvaK", "yvgN", "yvgO", "yvrE", "ywiE", "ywjC", "ywlB", "ywmE", "ywmF", "ywsB", "ywtG", "ywzA", "yxaB", "yxbG", "yxiS", "yxjI", "yxjJ", "yxkO", "yxnA", "yxzF", "yycD", "yczO", "ydaC", "yebE", "yebG", "yflD", "yflB", "yisP", "ipi", "yjlB", "ykzN", "spo0E", "yktC", "ykzC", "rbfA", "ytkC", "ytiB", "menC", "menE", "yuzH", "yvbG", "yvzE", "ywsA", "ywjA", "yydC", "yyzG", "yyzH", "yybO")


#cleanup the intermediate results
rm(gse6865_raw_df)
rm(na2)
rm(temp)
```
## Random Profiles

The random profiles are drawn from a [Gaussian process](https://en.wikipedia.org/wiki/Gaussian_process) with squared exponential kernel and zero mean function. To keep the profile strictly positive, it is transformed via $$f(x) = log(1 + e^x)$$

The source of the random profiles related functions can be found at https://github.com/cas-bioinf/genexpi/blob/master/rpackage/R/randomProfiles.R

```{r}

# Plot 10 random profiles with scale 5 and length 20
plotRandomProfiles(10, gse6865_raw_time, scale = 5, length = 20, main = "Sampled random profiles", ylab="Expression");
```

How do we identify the correct scale and length for the Gaussian process? Lets just find some that look similar enough to the sigB profile. After a bit of experimentation we chose scale = 7 and length = 20 minutes. Below is a sample of how those random trajectories look, compared to the sigB profile from the original data. (this is shown in the paper as Fig. 2)

```{r}
sigBRandomScale = 7
sigBRandomLength = 20
sigBIndex = which(rownames(gse6865_raw) == "sigB")
plotRandomProfiles(10,gse6865_raw_time, scale = sigBRandomScale, length = sigBRandomLength, trueTime = gse6865_raw_time, trueProfile = gse6865_raw[sigBIndex,], main = "Sampled random profiles and the sigB profile (dots)", ylab="expression")
```

## Genexpi in the sigB regulon

TODO - move function defs to package.
https://github.com/cas-bioinf/genexpi/blob/master/rpackage/R/evaluationTools.R

So lets take a sneak peek (just 3 random rounds) do we work with 6 degrees of freedom for the spline? (This will take a while to recompute).

```{r}
result6DF = evaluateRandomForRegulon(deviceSpecs, rounds = 3, rawProfiles = gse6865_raw,regulatorName =  "sigB", regulonNames = sigBRegulonNames, time = smoothTime, rawTime = gse6865_raw_time, randomScale = sigBRandomScale, randomLength = sigBRandomLength, errorDef = errorDef, splineDFs = 6)
```
```{r}
paste0("Proportion of true regulations discovered: ",result6DF$trueRatio)
paste0("Overall proportion of regulations by random profiles: ", result6DF$overallRandomRatio)
```

Now, let's test for various splines (we can't get less DFs than 3, 10 DFs do clearly overfit) and with more rounds.

```{r}
dfsToTest = 3:10
randomRounds = 50
genexpiStart = proc.time();
genexpiOptions(verbose = FALSE) # Hide unnecessary output
variousSplinesResult = testVariousSplines(deviceSpecs, rounds = randomRounds, rawProfiles = gse6865_raw, rawTime = gse6865_raw_time, targetTime = smoothTime, dfsToTest = dfsToTest, regulatorName = "sigB", regulonNames = sigBRegulonNames,randomScale = sigBRandomScale, randomLength = sigBRandomLength, errorDef = errorDef )

genexpiEnd = proc.time();
# Calculate time per round. For each spline (df) variant I have 1 non-random round in addition to the random roudns
genexpiTimePerRound = (genexpiEnd["elapsed"] - genexpiStart["elapsed"]) / (length(dfsToTest) * (randomRounds + 1))

printVariousSplinesResults(variousSplinesResult)
cat(paste0("Genexpi time per round: ", genexpiTimePerRound, " seconds\n"))

```

Let's plot histogram of the random results for 6DFs:
```{r}
hist(variousSplinesResult[[which(dfsToTest == 6)]]$result$randomRatios, main="", xlab = "Proportion predicted regulated by random profile.")
#TODO plot the highest scoring profiles
```

## Testing TD-Aracne in the sigB regulon

We cannot tell TD-Aracne to only test certain regulations. We tried running pairwise runs, but that turned out to be very expensive (random testing did not finish in a day, running )

When running against the full matrix, TD-Aracne is actually solving a much more complex problem than Genexpi.
Direct vs. downstream vs. connected
The sources for the evaluateTDAracne function and related code can be found at
https://github.com/cas-bioinf/genexpi/blob/master/rpackage/R/tdaracneBridge.R
```{r}
#Aracne requires uniform time intervals
aracneSmoothTimeIndices = c(1,11,21,31,41,51,61,71,81,91,101)
aracneRawTimeIndices = c(1,3,5,7:14)
#This was chosen as the minimal value that does not rise warnings
defaultAracneNumBins = 10

#Aracne is too slow to run too many rounds
randomRoundsAracne = 10
```


```{r}

aracneSplinedPairwiseStart =  proc.time();
aracneSplinedPairwiseResults = list()
for(i in 1:length(dfsToTest)) {
  aracneSplinedPairwiseResults[[i]] = evaluateTDAracnePairwise(
    title = paste("Spline_",dfsToTest[i]), rounds = randomRoundsAracne, profilesRaw = gse6865_raw, 
    time =  smoothTime[aracneSmoothTimeIndices], rawTime = gse6865_raw_time, splineDFs = dfsToTest[i],
    randomScale = sigBRandomScale, randomLength = sigBRandomLength,errorDef =errorDef, regulatorName = "sigB", regulonNames = sigBRegulonNames, numBins = defaultAracneNumBins)
}
aracneSplinedPairwiseEnd = proc.time();
aracneSplinedPairwiseTimePerRound = (aracneSplinedPairwiseEnd["elapsed"] - aracneSplinedPairwiseStart["elapsed"]) / ((randomRoundsAracne + 1) * length(dfsToTest));


aracneRawPairwiseStart =  proc.time();
aracneRawPairwiseResult = evaluateTDAracnePairwise(
  title = "Raw",rounds = randomRoundsAracne, profilesRaw = gse6865_raw[,aracneRawTimeIndices, drop=FALSE], 
  rawTime =  gse6865_raw_time[aracneRawTimeIndices], randomScale = sigBRandomScale, randomLength = sigBRandomLength,
  regulatorName = "sigB", regulonNames = sigBRegulonNames, errorDef =errorDef, numBins = defaultAracneNumBins,
  splineDFs = NULL, time = NULL)
aracneRawPairwiseEnd =  proc.time();
aracneRawPairwiseTimePerRound = (aracneRawPairwiseEnd["elapsed"] - aracneRawPairwiseStart["elapsed"]) / (randomRoundsAracne + 1);


numCoresUsed = detectCores()[1] - 1; #The script leaves one CPU idle to let you work while it computes

cat("==== Results for pairwise TD-ARACNE ====\n\n")
#Run on raw data is compared against the highest definition spline
printTDAracnePairwiseEvaluation("Raw data  ", aracneRawPairwiseResult, variousSplinesResult[[length(variousSplinesResult)]])
for(i in 1:length(dfsToTest)) {
  printTDAracnePairwiseEvaluation(paste0("Spline - ", dfsToTest[i], " DF"), aracneSplinedPairwiseResults[[i]], variousSplinesResult[[i]])
}
cat("\n")
cat(paste0("Raw time per round (on ", numCoresUsed," cores): ", aracneRawPairwiseTimePerRound, " seconds\n")) 
cat(paste0("Splined time per round (on ", numCoresUsed," cores): ", aracneSplinedPairwiseTimePerRound , " seconds\n"))

```

```{r}


aracneSplinedStart =  proc.time();
aracneSplinedResults = list()
for(i in 1:length(dfsToTest)) {
  aracneSplinedResults[[i]] = evaluateTDAracne(
    rounds = randomRoundsAracne, profilesRaw = gse6865_raw, time = smoothTime[aracneSmoothTimeIndices], 
    rawTime = gse6865_raw_time, splineDFs = dfsToTest[i],
    randomScale = sigBRandomScale, randomLength = sigBRandomLength,errorDef =errorDef, regulatorName = "sigB", regulonNames = sigBRegulonNames, numBins = defaultAracneNumBins)
}
aracneSplinedEnd = proc.time();
aracneSplinedTimePerRound = (aracneSplinedEnd["elapsed"] - aracneSplinedStart["elapsed"]) / ((randomRoundsAracne + 1) * length(dfsToTest));


aracneRawStart =  proc.time();
aracneRawResult = evaluateTDAracne(rounds = randomRoundsAracne, profileMatrix = gse6865_raw[,aracneRawTimeIndices, drop=FALSE], time =  gse6865_raw_time[aracneRawTimeIndices], scale = sigBRandomScale, length = sigBRandomLength,errorDef =errorDef, regulatorName = "sigB", regulonNames = sigBRegulonNames, numBins = defaultAracneNumBins)
aracneRawEnd =  proc.time();
aracneRawTimePerRound = (aracneRawEnd["elapsed"] - aracneRawStart["elapsed"]) / (randomRoundsAracne + 1);



numCoresUsed = detectCores()[1] - 1; #The script leaves one CPU idle to let you work while it computes
printTDAracneEvaluation("Raw data ", aracneRawResult)
cat(paste0("Raw time per round (on ", numCoresUsed," cores): ", aracneRawTimePerRound / 60, " minutes\n"))
for(i in 1:length(dfsToTest)) {
  printTDAracneEvaluation(paste0("Spline - ", dfsToTest[i], " DF"), aracneSplinedResults[[i]])
}
cat(paste0("Splined time per round (on ", numCoresUsed," cores): ", aracneSplinedTimePerRound / 60, " minutes\n"))

```
TODO: ARACNE a constant/const. synth genes.

# The SigR regulon of S. coelicolor
```{r}
#The data are included in the package
data(gse44415_raw)
gse44415_raw_time = as.numeric(colnames(gse44415_raw))
smoothTime_r = -30:330

sigRSco = "5216"
sigRRegulonSco = c("0169", "0170", "0197", "0200", "0204", "0213", "0216", "0219", "0220", "0494", "0498", "0961", "1814", "1815", "2088", "2154", "2155", "2223", "2359", "2389", "2562", "2687", "3124", "3209", "3350", "3906", "3907", "3909", "3945", "4094", "4232", "4649", "4653", "4654", "4655", "4659", "4661", "4701", "4702", "4703", "4704", "4705", "4706", "4708", "4709", "4710", "4711", "4712", "4713", "4714", "4715", "4716", "4717", "4718", "4719", "4720", "4721", "4725", "4726", "4728", "4729", "4855", "4901", "5012", "5295", "5366", "5368", "5369", "5370", "5371", "5372", "5373", "5374", "5624");

sigRRegulon_Kim = c("0569", "0570", "0882", "0884", "0885", "0917", "0973", "1084", "1085", "1142", "1238", "1425", "1426", "1513", "1598", "1599", "1600", "1618", "1619", "1643", "1644", "1645", "1646", "1647", "1648", "1758", "1869", "1919", "1920", "1936", "1937", "1938", "1939", "1958", "1995", "1996", "1997", "2124", "2154", "2161", "2162", "2194", "2254", "2310", "2331", "2481", "2537", "2538", "2539", "2595", "2617", "2618", "2619", "2634", "2635", "2642", "2643", "2763", "2816", "2849", "2910", "2911", "s02", "3083", "3091", "3162", "3187", "3206", "3207", "3295", "3296", "3373", "3403", "3442", "3449", "3450", "3451", "3509", "3764", "3765", "3766", "3767", "3889", "3890", "4039", "4040", "4109", "4203", "4204", "4205", "4297", "4298", "4299", "4418", "4419", "4420", "4561", "4770", "4771", "4797", "4833", "4834", "4835", "4956", "4966", "4967", "4968", "5042", "5065", "5163", "5178", "5187", "5188", "5216", "5217", "5284", "5285", "5357", "5359", "5360", "5361", "5465", "5490", "5545", "5552", "5705", "5706", "5707", "5708", "5709", "5754", "5796", "5820", "5864", "5865", "6061", "6126", "6127", "6423", "6551", "6759", "6760", "6761", "6762", "6763", "6764", "6765", "6766", "6767", "6768", "6769", "6770", "6771", "6775", "6776", "7140", "7631", "7632", "7784", "7785")

sigRRegulon_Kim_Strong = c("0570", "0882", "0884", "0885", "0917", "0973", "1084", "1085", "1142", "1238", "1425", "1426", "1513", "1600", "1619", "1648", "1758", "1869", "1920", "1936", "1958", "1995", "1997", "2124", "2154", "2161", "2162", "2194", "2254", "2310", "2331", "2481", "2537", "2595", "2619", "2634", "2635", "2642", "2643", "2763", "2816", "2849", "2911", "s02", "3083", "3091", "3162", "3187", "3206", "3207", "3296", "3373", "3403", "3442", "3450", "3451", "3509", "3765", "3766", "3767", "3890", "4039", "4040", "4109", "4203", "4204", "4297", "4419", "4420", "4561", "4770", "4797", "4835", "4956", "4966", "4967", "5042", "5065", "5163", "5178", "5187", "5188", "5216", "5284", "5285", "5357", "5359", "5465", "5490", "5545", "5552", "5705", "5754", "5796", "5820", "5864", "6061", "6126", "6127", "6423", "6551", "6759", "6775", "6776", "7140", "7631", "7632", "7784")
```

Once again, we try to find a setting for the profile generator to roughly match the sigR profile.

```{r}
sigRRandomScale = 3
sigRRandomLength = 30
sigRIndex = which(rownames(gse44415_raw) == sigRSco)
plotRandomProfiles(10,smoothTime_r, sigRRandomScale, sigRRandomLength, trueTime = gse44415_raw_time, trueProfile = gse44415_raw[sigRIndex,], main = "Sampled random profiles and the sigR profile (dots)", ylab="expression")
```

## Genexpi and SigR regulon
```{r}
dfsToTest_r = 3:11
genexpiStart_r = proc.time();
genexpiOptions(verbose = FALSE) # Hide unnecessary output
variousSplinesResult_r = testVariousSplines(deviceSpecs, rounds = randomRounds, rawProfiles = gse44415_raw[rownames(gse44415_raw) %in% c(sigRSco, sigRRegulonSco),], rawTime = gse44415_raw_time, targetTime = smoothTime_r, dfsToTest = dfsToTest_r, regulatorName = sigRSco, regulonNames = sigRRegulonSco,randomScale = sigRRandomScale, randomLength = sigRRandomLength, errorDef = errorDef )

genexpiEnd_r = proc.time();
# Calculate time per round. For each spline (df) variant I have 1 non-random round in addition to the random roudns
genexpiTimePerRound_r = (genexpiEnd_r["elapsed"] - genexpiStart_r["elapsed"]) / (length(dfsToTest) * (randomRounds + 1))

cat("\n=== Results for the sigR regulon ===\n")
printVariousSplinesResults(variousSplinesResult_r_kim_strong)

cat(paste0("Genexpi time per round: ", genexpiTimePerRound_r, " seconds\n"))

```

```{r}
genexpiStart_r_kim = proc.time();
genexpiOptions(verbose = FALSE) # Hide unnecessary output
variousSplinesResult_r_kim = testVariousSplines(deviceSpecs, rounds = randomRounds, rawProfiles = gse44415_raw[rownames(gse44415_raw) %in% c(sigRSco, sigRRegulon_Kim),], rawTime = gse44415_raw_time, targetTime = smoothTime_r, dfsToTest = dfsToTest2, regulatorName = sigRSco, regulonNames = sigRRegulon_Kim,randomScale = sigRRandomScale, randomLength = sigRRandomLength, errorDef = errorDef )

genexpiEnd_r_kim = proc.time();
# Calculate time per round. For each spline (df) variant I have 1 non-random round in addition to the random roudns
genexpiTimePerRound_r_kim = (genexpiEnd_r_kim["elapsed"] - genexpiStart_r_kim["elapsed"]) / (length(dfsToTest) * (randomRounds + 1))

cat("=== Results for the whole sigR regulon of Kim et al. ===\n")
printVariousSplinesResults(variousSplinesResult_r_kim)
cat(paste0("Genexpi time per round: ", genexpiTimePerRound_r_kim, " seconds\n"))


genexpiStart_r_kim_strong = proc.time();
genexpiOptions(verbose = FALSE) # Hide unnecessary output
variousSplinesResult_r_kim_strong = testVariousSplines(deviceSpecs, rounds = randomRounds, rawProfiles = gse44415_raw[rownames(gse44415_raw) %in% c(sigRSco, sigRRegulon_Kim_Strong),], rawTime = gse44415_raw_time, targetTime = smoothTime_r, dfsToTest = dfsToTest2, regulatorName = sigRSco, regulonNames = sigRRegulon_Kim_Strong, randomScale = sigRRandomScale, randomLength = sigRRandomLength, errorDef = errorDef )

genexpiEnd_r_kim_strong = proc.time();
# Calculate time per round. For each spline (df) variant I have 1 non-random round in addition to the random roudns
genexpiTimePerRound_r_kim_strong = (genexpiEnd_r_kim_strong["elapsed"] - genexpiStart_r_kim_strong["elapsed"]) / (length(dfsToTest) * (randomRounds + 1))

cat("\n=== Results for the 'STRONG' subpart of sigR regulon of Kim et al. ===\n")
printVariousSplinesResults(variousSplinesResult_r_kim_strong)

cat(paste0("Genexpi time per round: ", genexpiTimePerRound_r_kim_strong, " seconds\n"))
```

## TD-Aracne and sigRRegulon
```{r}
aracneSplinedPairwiseStart_r =  proc.time();
aracneSplinedPairwiseResults_r = list()
aracneSmoothTimeIndices_r = seq(1,361, by = 30)

for(i in 1:length(dfsToTest_r)) {
  aracneSplinedPairwiseResults_r[[i]] = evaluateTDAracnePairwise(
    title = paste("Spline_",dfsToTest[i]), rounds = randomRoundsAracne, profilesRaw = gse44415_raw, 
    time =  smoothTime[aracneSmoothTimeIndices], rawTime = gse44415_raw_time, splineDFs = dfsToTest_r[i],
    randomScale = sigRRandomScale, randomLength = sigRRandomLength,errorDef =errorDef, regulatorName = sigRSco, regulonNames = sigRRegulonSco, numBins = defaultAracneNumBins)
}
aracneSplinedPairwiseEnd_r = proc.time();
aracneSplinedPairwiseTimePerRound_r = (aracneSplinedPairwiseEnd_r["elapsed"] - aracneSplinedPairwiseStart_r["elapsed"]) / ((randomRoundsAracne + 1) * length(dfsToTest_r));


aracneRawPairwiseStart_r =  proc.time();
aracneRawPairwiseResult_r = evaluateTDAracnePairwise(
  title = "Raw",rounds = randomRoundsAracne, profilesRaw = gse44415_raw, 
  rawTime =  gse6865_raw_time[aracneRawTimeIndices], randomScale = sigRRandomScale, randomLength = sigRRandomLength,
  regulatorName = sigRSco, regulonNames = sigRRegulonSco, errorDef =errorDef, numBins = defaultAracneNumBins,
  splineDFs = NULL, time = NULL)
aracneRawPairwiseEnd_r =  proc.time();
aracneRawPairwiseTimePerRound_r = (aracneRawPairwiseEnd_r["elapsed"] - aracneRawPairwiseStart_r["elapsed"]) / (randomRoundsAracne + 1);


numCoresUsed = detectCores()[1] - 1; #The script leaves one CPU idle to let you work while it computes

cat("==== Results for pairwise TD-ARACNE ====\n\n")
#Run on raw data is compared against the highest definition spline
printTDAracnePairwiseEvaluation("Raw data  ", aracneRawPairwiseResult_r, variousSplinesResult_r[[length(variousSplinesResult_r)]])
for(i in 1:length(dfsToTest_r)) {
  printTDAracnePairwiseEvaluation(paste0("Spline - ", dfsToTest_r[i], " DF"), aracneSplinedPairwiseResults_r[[i]], variousSplinesResult_r[[i]])
}
cat("\n")
cat(paste0("Raw time per round (on ", numCoresUsed," cores): ", aracneRawPairwiseTimePerRound_r, " seconds\n")) 
cat(paste0("Splined time per round (on ", numCoresUsed," cores): ", aracneSplinedPairwiseTimePerRound_r , " seconds\n"))
```

