---
title: "Evaluating Sacharomyces cell cycle genes"
output: html_notebook
---
# Setting up

```{r setup}
  options(java.parameters = "-Xmx2048m")
```

The code below should install all required packages for the evaluation.

```{r eval=FALSE}
  # Helper function to install packages if they are not available
  pkgTest <- function(x)
  {
    if (!require(x,character.only = TRUE))
    {
      install.packages(x,dep=TRUE)
        if(!require(x,character.only = TRUE)) stop("Package not found")
    }
  }

  source("https://bioconductor.org/biocLite.R")
  biocLite(c("Biobase","TDARACNE", "RBGL"))

  pkgTest("splines")
  pkgTest("foreach")
  pkgTest("doParallel")
  pkgTest("rJava")
  pkgTest("tidyverse")
  
  pkgTest("devtools")
  install_github("cas-bioinf/genexpi", ref = "bmc-release", subdir="rpackage")

```

Load the packages

```{r results='hide'}
  library(Biobase)
  library(TDARACNE)
  library(RBGL)
  
  library(splines)
  library(foreach)
  library(doParallel)
  library(rJava)
  library(tidyverse)
  library(rGenexpi)
```

## Download the data from Yeastract and GEO datasets

```{r}
yeastract_file <- 'yeastract.tsv.gz'
if(!file.exists(yeastract_file)) {
  download.file("http://www.yeastract.com/download/RegulationTwoColumnTable_Documented_2013927.tsv.gz", yeastract_file)
}

yeastract_df <- read.delim(gzfile(yeastract_file),sep = ";",header = FALSE, col.names = c("regulator","target"))
```



```{r}
gds38_file <- 'gds38.soft.gz';
if(!file.exists(gds38_file)) {
  download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/datasets/GDSnnn/GDS38/soft/GDS38_full.soft.gz", gds38_file)
}

#Intermediate data frame representation
gds38_raw_df = read.delim(gzfile(gds38_file), skip = 149, comment.char = "!", na.strings = c("null","")) 

```

## Data processing

```{r}
#SWI4/6, MBP1 a re suspect as they act in complexes (BTW good test case for the cooperative model)

regulators <- c("FKH1", "FKH2","MCM1","MBP1","SWI4","SWI6","NDD1", "ACE2")
yeastract_filtered <- yeastract_df%>% filter(regulator %in% regulators) %>%
  filter(target %in% gds38_raw_df$IDENTIFIER) %>%
  group_by(regulator)

all_targets <- as.character(yeastract_filtered$target)
genes_of_interest <- unique(as.character(c(regulators, all_targets)))

gds38_raw <- aggregate(gds38_raw_df[,3:18], gds38_raw_df["IDENTIFIER"], mean, na.rm = TRUE) %>%
  column_to_rownames("IDENTIFIER") %>%
  as.matrix() %>% exp()
  
gds38_raw_filtered <- gds38_raw[genes_of_interest,]

time_raw <- seq(0,by = 7, length.out = 16)
time_smooth <- 0:max(time_raw)
```

## Splining the profiles + inspecting results
```{r}
splineDF = 6
gds38_smooth_filtered <- splineProfileMatrix(gds38_raw_filtered, time_raw, time_smooth, df = splineDF, intercept = TRUE)
inspectSmoothing(time_raw, gds38_raw_filtered,time_smooth, gds38_smooth_filtered, c("ACE2","FKH2","SWI4"))

```
# The evaluation itself

```{r}
errorDef <- defaultErrorDef()
errorDef$relative <- 0.1
errorDef$minimal <- 0.2
deviceSpecs <- getDeviceSpecs(deviceType = "processor")
minFitQuality <- 0.8
randomRounds <- 20
```

## Filter out constant synthesis
```{r}
regulators_logical <- rownames(gds38_smooth_filtered) %in% regulators 
constantProfiles = testConstant(gds38_smooth_filtered, errorDef)
profilesToTestConstantSynthesisIndices = which(!constantProfiles & !regulators_logical)

constantSynthesisResults = computeConstantSynthesis(deviceSpecs, gds38_smooth_filtered, tasks = profilesToTestConstantSynthesisIndices);

constantSynthesisProfiles = testConstantSynthesis(constantSynthesisResults, errorDef, minFitQuality);

profilesToTest_logical <- regulators_logical | (!constantProfiles & !constantSynthesisProfiles)

gds38_raw_to_test <- gds38_raw_filtered[profilesToTest_logical,]
gds38_smooth_to_test <- gds38_smooth_filtered[profilesToTest_logical,]

targets_to_test <- rownames(gds38_smooth_filtered)[(!constantProfiles & !constantSynthesisProfiles)]
```

## Test random profile generation

Here the dots are actual measured profiles of several regulators, while the lines are randomly generated profiles.

```{r}
randomScale = 1.6
randomLength = 10
#See ACE2
plotRandomProfiles(20, time_raw, scale = randomScale, length = randomLength,trueTime = time_raw,  trueProfile = gds38_raw["ACE2",])
#See FKH1
plotRandomProfiles(20, time_raw, scale = randomScale, length = randomLength,trueTime = time_raw,  trueProfile = gds38_raw["FKH1",])
#See SWI4
plotRandomProfiles(20, time_raw, scale = randomScale, length =                randomLength,trueTime = time_raw,  trueProfile = gds38_raw["SWI4",])
```

## Subsample the regulons

```{r}
results_genexpi <- list()
results_aracne <- list()

set.seed(21345678)
regulons_to_test = list()
for(regulator_to_test in regulators) {
  regulon_to_test_df <- yeastract_filtered %>% filter(regulator == regulator_to_test & target %in% targets_to_test)
  if(nrow(regulon_to_test_df) > 30) {
    regulon_to_test_df <- regulon_to_test_df %>% sample_n(30)
  }
  regulons_to_test[[regulator_to_test]] <- regulon_to_test_df$target %>% as.character()
}

```

## Execute Genexpi and TD-Aracne
```{r results='hide'}

for(regulator_to_test in regulators) {
  cat("======", regulator_to_test, "=====\n")
  regulon_to_test <- regulons_to_test[[regulator_to_test]]
  
  #This is tu support restarts of the computation
  if(is.null(results_genexpi[[regulator_to_test]])) {
    evaluation_res <- evaluateRandomForRegulon(deviceSpecs = deviceSpecs,rawProfiles = gds38_raw_to_test, rounds = randomRounds, regulatorName = regulator_to_test, regulonNames = regulon_to_test, checkConstantSynthesis = FALSE, time = time_smooth, rawTime = time_raw, randomScale = randomScale, randomLength = randomLength, splineDFs = splineDF, splineIntercept = TRUE, errorDef = errorDef)
    results_genexpi[[regulator_to_test]] <- evaluation_res
    save.image("evaluation_sacharomyces.RData")
  }
}

for(regulator_to_test in regulators) {
  cat("====== Aracne:", regulator_to_test, "=====\n")
  regulon_to_test <- regulons_to_test[[regulator_to_test]]
  #This is tu support restarts of the computation
  if(is.null(results_aracne[[regulator_to_test]])) {
    res_aracne <- evaluateTDAracne(randomRounds, gds38_raw, time_raw, c(splineDF), time_smooth, randomScale, randomLength, regulatorName = regulator_to_test, regulonNames = regulon_to_test, errorDef = errorDef, numBins = 10)
    results_aracne[[regulator_to_test]] <- res_aracne
    save.image("evaluation_sacharomyces.RData")
  }
  
}
```

## Output the results

For the publication, we chose the better result for TD-Aracne (either downstream or connected) separately for each regulator.

```{r}
results_formatted <- list()
step = 1
for(regulator in regulators) {
  if(!is.null(results_genexpi[[regulator]])) {
    res_genexpi <- results_genexpi[[regulator]]
    results_formatted[[step]] <- data.frame(regulator = regulator, type = "Genexpi", true = res_genexpi$trueRatio, random = mean(res_genexpi$randomRatios))
    step <- step + 1
  }
  
  if(!is.null(results_aracne[[regulator]])) {
    res_aracne <- results_aracne[[regulator]]
    results_formatted[[step]] <- data.frame(regulator = regulator, type = "TD-Aracne-Downstream", true = res_aracne$trueRatioDownstream, random = res_aracne$overallRandomRatioDownstream)
    step <- step + 1
    results_formatted[[step]] <- data.frame(regulator = regulator, type = "TD-Aracne-Connected", true = res_aracne$trueRatioConnected, random = res_aracne$overallRandomRatioConnected)
    step <- step + 1
  }
}

results_formatted <- do.call(rbind, results_formatted)
results_formatted
```

